Next Generation Planning - Structuring and Sharing Environmental Drivers Data for the St. Lawrence System

### 2019-05-01 eDriversMS-v2.Rmd
latest modifications to manuscript, including frontiers formatting and supervisor comments

---
title: Next Generation Planning - Structuring and Sharing Environmental Drivers Data for the St. Lawrence
author: David Beauchesne^1,2,\*^, Rémi M. Daigle^2^, Steve Vissault^3^, Dominique Gravel^3^, Andréane Bastien^4^, Simon Bélanger^5^, Pascal Bernatchez^5^, Marjolaine Blais ^6^, Hugo Bourdages^6^, Clément Chion^7^, Peter S. Galbraith^6^, Benjamin Halpern^8,9^, Camille Lavoie^2^, Christopher W. McKindsey^6^, Alfonso Mucci^10^, Simon Pineault^11^, Michel Starr^6^, Anne-Sophie Ste-Marie^4^, Philippe Archambault^2^
date: 2019-04-23
fontsize: 12pt
font-family: Times New Roman
output:
  pdf_document:
    toc: false
    number_sections: true
  word_document:
    toc: false
header-includes:
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{listings}
   - \usepackage[svgnames]{xcolor}
   # - \usepackage{titling}
   # - \pretitle{\begin{flushleft}}
   # - \posttitle{\end{flushleft}}
   - \usepackage{lscape}
   - \newcommand{\blandscape}{\begin{landscape}}
   - \newcommand{\elandscape}{\end{landscape}}
   - \usepackage{bbm}
   - \usepackage{float}
relativeurls: true
bibliography: eDriversMS.bib
csl: frontiers.csl
link-citations: yes
---

<!--
cd phdproj/eDrivers/eDriversMS/
rmarkdown::render('./eDriversMS-v2.rmd', 'pdf_document')
rmarkdown::render('./eDriversMS.rmd', 'word_document')
-->

@@

**Number of figures**: 5 \newline
**Number of tables**: 1 \newline

@@

<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->
# Abstract
<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->
<!-- 350 words -->

The St. Lawrence System is a vast and complex social-ecological system providing
a wealth of ecosystem services sustaining numerous economic sectors. Related
human pressures combine with climate driven environmental changes to produce
intricate exposure regimes of overlapping and potentially interacting
environmental drivers (*e.g.* hypoxia and fisheries) threatening ecosystems.
There is a critical need to systematically characterize the distribution,
intensity and overlap between drivers to support integrative initiatives such
as ecosystem-based management. While portals providing knowledge on individual
drivers exist, platforms collating comparable and interoperable knowledges on
multiple drivers remain conspicuously missing. This paper presents two distinct,
but related objectives. The first is to characterize the distribution and
intensity of drivers in the St. Lawrence System. The second is to launch
*eDrivers*, an open knowledge platform gathering experts committed to
structuring, standardizing and sharing knowledges on drivers in support of
science and management. We gathered data on 22 coastal, climate, fisheries and
marine traffic drivers through collaborations, existing environmental
initiatives and open data portals. We show that few areas of the St. Lawrence
are free of cumulative exposure. The Estuary, the Anticosti Gyre and coastal
areas are particularly exposed, especially in the vicinity of urban centers.
We identified 6 areas of distinct cumulative exposure regime that show that
certain drivers typically co-occur in different regions of the St. Lawrence and
that coastal areas are exposed to all driver types. Of particular concern are
two threat complexes capturing most exposure hotspots that show the convergence
of contrasting exposure regimes at the head of the Laurentian Channel. These
observations are destined to improve as *eDrivers* evolves through time to
address knowledge gaps and refine current driver layers. In an effort to share
the knowledge acquired and to ensure the lasting relevance of the description
of drivers presented in this manuscript, *eDrivers* was built on a series of
guiding principles upholding existing data management and open science
standards. Ultimately, we believe that *eDrivers* represents a much needed
solution that could radically influence broad scale research and management
practices by increasing data accessibility and interoperability and by
increasing research and decision-making efficiency.

@@

The St. Lawrence System, formed by one of the largest estuaries in the world and
a vast interior sea, is a complex social-ecological system characterized by
highly variable environmental conditions and oceanographic processes, both in
space and time [@el-sabh1990a; @white1997; @dufour2007]. It thus offers a unique
and heterogeneous array of habitats suited for the establishment of diverse and
productive ecological communities [@savenkoff2000]. As a result, the St.
Lawrence System provides a wealth of ecosystem services that have historically
and contemporarily benefited the Canadian economy. It sustains a rich fisheries
industry targeting more than 50 species, serves as the gateway to eastern
North-America by granting access to more than 40 ports and the most densely
populated Canadian region, hosts a booming tourism industry and an expanding
aquaculture production, fosters emerging activities and boasts a yet untapped
hydrocarbon potential [@beauchesne2016; @schloss2017 @archambault2017]. With
major investments recently made and more forthcoming in economic and
infrastructure development and research [*e.g.* @governmentofquebec2015; @rqm2018], an
intensification of the human footprint is expected in the St. Lawrence System.

As elsewhere in the world [see @halpern2015a], this intensifying human footprint
will likely result in increasingly intricate environmental exposure regimes,
*i.e.* suites of overlapping and potentially interacting environmental drivers
threatening ecosystems, habitats or ecological communities. Drivers, often
referred to as stressors or pressures, are any externalities that affect
environmental processes and disturb natural systems. Drivers may originate from
natural or human-induced biophysical processes (*e.g.* sea surface temperature
anomalies and hypoxia) or directly from anthropogenic activities (*e.g.*
fisheries and marine pollution). The potential for complex interactions between
drivers is the largest uncertainty when studying or predicting environmental
change [@darling2008; @cote2016]. The effects of multiple drivers can combine
non-linearly and result in effects that are greater (synergistic effect) or
lower (antagonistic effect) than the sum of individual effects
[@crain2008; @darling2008; @cote2016].  <!-- For example...??? -->

The uncertainty associated with complex driver interactions must therefore be
taken into account when investigating environmental impacts [@cote2016], yet
most research on driver effects in marine environments remains overwhelmingly
focused on single driver assessments [@obrien2019]. Arising from the expected
increase in environmental exposure and the experiences of past and
current ecological tragedies such as the collapse of cod fisheries
[@frank2005; @dempsey2018] and the decline of the beluga and right whale <!-- right whale ref -->
populations [@plourde2014] is a need to characterize the distribution,
intensity and overlap between drivers in the St. Lawrence System. This will
provide critical information on areas most exposed to cumulative drivers and on
the interaction potential of drivers in the St. Lawrence. It is also a necessary
step towards the holistic and integrated management of the St. Lawrence System.

Gathering environmental data for large scale, systematic initiatives can,
however, be a very challenging and time consuming - not to say painful - process.
On one hand, there is an overwhelming and expanding wealth of data available.
Such information overload may inhibit our ability to make decisions based on
scientific information, promote massive effort duplication, disproportionately
appropriate research funds to certain sectors, and obscure knowledge gaps amid
a sea of information [@eppler2004]. On the other hand, crucial data are lacking
and remain largely unavailable or inaccessible for a variety of reasons,
including proprietary rights, lack of organizational time, capacity and
training, and in some rare cases unwillingness to share, curtailing our ability
for appropriate decision-making.

The current digital infrastructure is highly decentralized and the data
management and sharing practices are highly heterogeneous, preventing us from
maximizing benefits from research investments [@wilkinson2016].
Yet there now exists multiple initiatives that address this issue by assembling,
organizing and sharing vast arrays of environmental data.
Biotic data can be accessed through web portals such as the Ocean Biogeographic
Information System [OBIS; @obis2018], the Global Biodiversity Information Facility
[GBIF; @gbif2018], the Global Biotic Interactions platform [GloBI; @poelen2014],
and the World Register of Marine Species [WoRMS; @wormseditorialboard2017]. Abiotic data can
also be accessed through WorldClim [@hijmans2005], Bio-ORACLE
[@tyberghein2012], and MARSPEC [@sbrocco2013]. Initiatives focused on sharing
environmental data for specific areas also exist, such as the St. Lawrence
Global Observatory (SLGO[^slgo]), the European Marine Observation
and Data Network (EMODnet[^emod]) and the U.S. Integrated Ocean
Observing System (IOOS[^ioos]). Essential environmental
parameters are also organized, coordinated and acquired through global
initiatives like the Group on Earth Observations Biodiversity Observation
Network [GEO BON; @scholes2012], the Census of Marine Life [CoML; @coml2010]
and the Global Ocean Observing System (GOOS[^goos]).

[^slgo]: https://ogsl.ca/en
[^emod]: http://www.emodnet.eu/
[^ioos]: https://ioos.noaa.gov/
[^goos]: http://www.goosocean.org

However, equivalent platforms for environmental drivers have largely focused on
single drivers (e.g. Global Fishing Watch) and platforms collating
data and knowledge on multiple drivers in a comparable and interoperable way
remain conspicuously missing [but see @halpern2015]. This is in spite of
integerated management and assessment approaches needing efficient data reporting,
standardized data management practices and tools tailored to the study of the
effects of multiple drivers [@dafforn2016; @stock2018]. An additional objective
thus emerged in the process of addressing our initial goals: sharing the
knowledge gathered through the description of drivers in the St.
Lawrence.

This manuscript thus has two distinct, but related, goals. The first
is to characterize the distribution and intensity of drivers in the St. Lawrence
System in order to: 1) identify areas of high cumulative exposure,
2) characterize areas with similar cumulative exposure regimes, dubbed threat
complexes [@bowler2019] and 3) identify drivers that are likely to interact
in the St. Lawrence. The second goal is to present how we are sharing the
knowledge acquired by launching an open knowledge platform, *eDrivers*, and use
this paper as a call for collaboration to any person, group or organization that
might be interested in contributing to this initiative.

@@

@@

The list of drivers for which we sought data was informed by a global cumulative
impact assessment initiative [@halpern2008b; @halpern2015a], regional holistic
evaluations of the state of the St. Lawrence [@dufour2007; @benoit2012], and
communications with regional experts. Through the data gathering process, we
developed and continue to develop collaborations with regional experts and data
holders. We also use global data from the global cumulative impact assessment
initiative [@halpern2008b; @halpern2015a] available from the National Center
for Ecological Analysis and Synthesis (NCEAS) online data repository
[Table 1; @halpern2015]. We selected global data that were unavailable at the
regional scale and that were available at a resolution adequate for use at the
scale of the St. Lawrence (*e.g.* marine pollution).

We were able to characterize the intensity and distribution of 22 drivers
(Table 1;) and we are actively working on updating or developing additional
driver layers. Drivers incorporated in the analyses are varied in origin, *i.e.* from
terrestrial (*e.g.* nutrient input) to marine (*e.g.* shipping), and from
large scale biophysical processes (*e.g.* temperature anomalies) to localized
anthropogenic activities (*e.g.* fisheries). Drivers were divided into 4 groups:
coastal, climate, fisheries and marine traffic (Table 1). All data layers and
methodologies are described in the supplementary materials.

Drivers with non-normal frequency distributions were log-transformed (Figure S1)
and all drivers were scaled between 0 and 1 to allow driver comparisons. The
99th quantile of individual driver distribution was used as the upper bound for
scaling to control for extreme values and produce maps of individual drivers
(Figure S2). All drivers were embedded in a regular grid composed of 245604
$1 km^2$ exagonal cells to construct the integrated dataset used for the
analyses.

@@

Areas with high cumulative exposure (objective 1) were identified by evaluating
the cumulative footprint of combined drivers and by identifying hotspots of
cumulative footprint. Cumulative footprint ($F$) was defined as the sum of
the scaled intensity of all drivers in each grid cell:

@@

where $x$ is a grid cell, $i$ is a driver and $D$ is the scaled intensity of
driver $i$.

Cumulative hotspots ($H$) were defined as the number of drivers in each grid
cell with scaled intensity contained over their respective 80th percentile:

@@

The distribution of driver interactions was investigated through the spatial
overlap of combinations of drivers using the cumulative footprint ($F$) equation
with pairs of drivers. The intensity at which pairs of drivers co-occur was
evaluated using a two-dimensional kernel density estimate. As there are 231
pairwise combinations between 22 drivers, we focus on a single example using
hypoxia and demersal destructive fisheries, two drivers known to occur mainly
in deeper areas of the St. Lawrence and, hence, an interaction between the
effects of the two drivers could be anticipated. Note, however, that *eDrivers*
will provide the capacity to compare any combinations of drivers.

@@

Natural systems are likely to host multiple overlapping drivers. The
dimensionality of the problem can quickly rise as additional drivers are
considered. In order to decrease the dimensionality of the integrated dataset,
we identify threat complexes, *i.e.* regions with similar cumulative exposure
regimes, using a clustering approach [*e.g.* see @bowler2019].

@@

Threat complexes were identified using a partional *k-medoids* clustering
algorithm, CLARA [CLustering for Large Applications; @kaufman1990], which was
designed for large datasets. The CLARA algorithm uses the PAM (Partition Around
Medoids) algorithm on a sample from the original dataset to identify a set of
$k$ objects that are representative of all other objects, *i.e.* medoids and
that are central to the cluster they represent. The goal of the algorithm is to
iteratively minimize intra-cluster dissimilarity. Iterations are compared on the
basis of the average dissimilarity between cluster objects and representative
medoid to select the optimal set of $k$ medoids that minimizes average
dissimilarity. We used 100 iterations using samples of 10000 observations
(*i.e.* ~5% of observations) to identify clusters. Analyses were performed
using the *cluster* R package [@maechler2018].

Partitional clustering algorithms require a user-defined number of clusters.
Values of $k$ ranging from 2 to 10 were tested and validated by selecting
the number of clusters that maximized the average silhouette width [@kaufman1990]
and minimized the total within-cluster sum of squares (Figure S3).

We also validated the clustering by comparing *k-medoids* clustering with
*k-means* clustering with the *Lloyd* algorithm [@lloyd1982]. The *k-means*
approach is similar to the *k-medoids*, but identifies observations belonging
to a cluster iteratively by minimizing the mean intra-cluster squared distance
until it converges to an optimal solution. We used 25 random sets and set a
maximum of 1000 iterations for the analysis. Analyses were performed using
the *stats* R package [@rcoreteam2018]. We used the same validation procedure
to select the optimal number of clusters $k$ than with the *k-medoids*
clustering (Figure S3).

While *k-means* algorithms are more  efficient since they do not compute
pairwise dissimilarities, it is more sensitive to outliers through the use of
the mean rather than a centroid. We therefore favored the use of the *k-medoids*
algorithm, but used the *k-means* to validate clusters.

@@

The difference between clusters was explored by measuring the total inter-cluster
dissimilarity and the contribution of each driver to the total inter-cluster
dissimilarity using a similarity percentage analysis (SIMPER) with Bray-Curtis
dissimilarity [Figure S4; @clarke1993]. As the drivers dataset is too
large, we used a bootstrap procedure for the SIMPER analysis, randomly selecting
5% of each cluster to run the analysis and repeating the process over 300
iterations. We also compared the mean intensity of each driver within each
cluster to better capture the inter-cluster dissimilarity. Analyses were
performed using the *vegan* R package [@oksanen2018].

@@

Intra-cluster similarity was evaluated using the Bray-Curtis similarity index
(Figure S5). As with the inter-cluster dissimilarity, we used a
bootstrap procedure for the intra-cluster similarity, randomly selecting 5% of
each cluster observation to run the analysis and repeating the process over 300
iterations. We however did not use the bootstraping procedure for clusters with
less than 10000 observations since computation time was manageable.

<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->
# Results and discussion
<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->

@@

Apart from the northeastern Gulf, the cumulative footprint of drivers is
ubiquitous in the St. Lawrence (Figure \ref{footprint}). Cumulative exposure is
generally highest along the coast (Figure \ref{footprint}), with hotposts
located in the vicinity of coastal cities (Figure \ref{hotspot}). In general,
offshore areas are less exposed to cumulative drivers, with the Estuary and the
Anticosti Gyre being notable exceptions (Figures \ref{footprint} and
\ref{hotspot}). This is not to say that offshore areas are free of exposure,
as most of the St. Lawrence is exposed to multiple overlapping drivers
(Figures \ref{footprint} and \ref{hotspot}). For example, it is worthy to note
high cumulative footprint observed at the heads of the Anticosti and Esquiman
Channels (Figure \ref{footprint}).

These results are consistent with observations elsewhere in the world, where
cumulative driver exposure conspicuously arises from and markedly intensifies
close to coastal cities and at the mouth of rivers draining highly populated
areas [e.g. @halpern2015a; @feist2016; @mach2017; @stock2018]. These are areas
where human activities (*e.g.* coastal development and shipping) and footprint
(*e.g.* pollution runoff) are the most intense [@feist2016], and on which is
overlaid a background of natural disturbances [@micheli2016]. They are also the
areas in which the most dramatic increases in exposure are expected, with
populations increasing more rapidly along the coast than inland [@feist2016].
In the St. Lawrence, large coastal cities are mostly located along the Estuary
and the southwestern Gulf, while the northeastern Gulf is largely uninhabited
or home to small coastal communities.

As for offshore exposure, the Estuary, along with the St. Lawrence River,
provide access to and serve as the primary drainage outflow of the Great Lakes
Basin, the most densely populated region in Canada [@statisticscanada2017]. Most marine
traffic thus converges to the Estuary.

@@

Box 1. Code snippet demonstrating how to use the *eDrivers* to reproduce
figure \ref{kernel} in R.

>>>>

---
title: Next Generation Planning - Structuring and Sharing Environmental Drivers Data for the St. Lawrence System
author: David Beauchesne^1,2,\*^, Rémi M. Daigle^2^, Steve Vissault^3^, Dominique Gravel^3^, Andréane Bastien^4^, Simon Bélanger^5^, Pascal Bernatchez^5^, Marjolaine Blais ^6^, Hugo Bourdages^6^, Clément Chion^7^, Peter S. Galbraith^6^, Benjamin Halpern^8,9^, Camille Lavoie^2^, Christopher W. McKindsey^6^, Alfonso Mucci^10^, Simon Pineault^11^, Michel Starr^6^, Anne-Sophie Ste-Marie^4^, Philippe Archambault^2^
fontsize: 12pt
font-family: Times New Roman
output:
  pdf_document:
    toc: false
    number_sections: true
  word_document:
    toc: false
header-includes:
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{listings}
   - \usepackage[svgnames]{xcolor}
   # - \usepackage{titling}
   # - \pretitle{\begin{flushleft}}
   # - \posttitle{\end{flushleft}}
   - \usepackage{lscape}
   - \newcommand{\blandscape}{\begin{landscape}}
   - \newcommand{\elandscape}{\end{landscape}}
   - \usepackage{bbm}
   - \usepackage{float}
relativeurls: true
bibliography: eDriversMS.bib
csl: frontiers.csl
link-citations: yes
---

<!--
cd phdproj/eDrivers/eDriversMS/
rmarkdown::render('./eDriversMS-v2.rmd', 'pdf_document')
rmarkdown::render('./eDriversMS-v2.rmd', 'word_document')
-->

@@

**Number of figures**: 6 \newline
**Number of tables**: 1 \newline

@@

<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->
# Abstract
<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->

The St. Lawrence is a vast and complex socio-ecological system providing a
wealth of services sustaining numerous economic sectors. These ecosystems are
subject to significant human pressures that overlap and potentially interact
with climate driven environmental changes. Our objective in this paper is to
systematically characterize the distribution and intensity of drivers in the
St. Lawrence System. To do so, we launch *eDrivers*, an open knowledge
platform gathering experts committed to structuring, standardizing and sharing
knowledge on drivers in support of science and management. We gathered data on
22 coastal, climate, fisheries and marine traffic drivers through
collaborations, existing environmental initiatives and open data portals. We
show that few areas of the St. Lawrence are free of cumulative exposure. The
Estuary, the Anticosti Gyre and coastal areas are particularly exposed,
especially in the vicinity of urban centers. We identified 6 areas of distinct
cumulative exposure regime that show that certain drivers typically co-occur
in different regions of the St. Lawrence and that coastal areas are exposed to
all driver types. Of particular concern are two threat complexes capturing
most exposure hotspots that show the convergence of contrasting exposure
regimes at the head of the Laurentian Channel. *eDrivers* was built on a
series of guiding principles upholding existing data management and open
science standards. We therefore expect it to evolve through time to address
knowledge gaps and refine current driver layers. Ultimately, we believe that
*eDrivers* represents a much needed solution that could radically influence
broad scale research and management practices by increasing knowledge accessibility
and interoperability.

@@

The St. Lawrence System, formed by one of the largest estuaries in the world,
the St. Lawrence Estuary, and a vast interior sea, the Gulf of St. Lawrence, is
a complex social-ecological system characterized by highly variable
environmental conditions and oceanographic processes
[@el-sabh1990a; @white1997; @dufour2007]. It constitutes a unique
and heterogeneous array of habitats suited for the establishment of diverse and
productive ecological communities [@savenkoff2000]. As a result, the St.
Lawrence System provides a wealth of ecosystem services that have historically
and contemporarily benefited the Canadian economy. It sustains a rich fisheries
industry targeting more than 50 species, serves as the gateway to eastern
North-America by granting access to more than 40 ports and the most densely
populated Canadian region, hosts a booming tourism industry and an expanding
aquaculture production, fosters emerging activities and boasts a yet untapped
hydrocarbon potential [@beauchesne2016; @schloss2017; @archambault2017]. With
major investments recently made and more forthcoming in economic and
infrastructure development and research [*e.g.* @governmentofquebec2015; @rqm2018],
an intensification of the human footprint is likely in the St. Lawrence System.

As elsewhere in the world [see @halpern2015a], this intensifying human footprint
will likely result in increasingly intricate environmental exposure regimes,
*i.e.* suites of overlapping and potentially interacting environmental drivers
threatening ecosystems, habitats or ecological communities. Drivers, often
referred to as stressors or pressures, are any externalities that affect
environmental processes and disturb natural systems. Drivers may originate from
natural or human-induced biophysical processes (*e.g.* sea surface temperature
anomalies and hypoxia) or directly from anthropogenic activities (*e.g.*
fisheries and marine pollution). The potential for complex interactions between
drivers is the largest uncertainty when studying or predicting environmental
impacts [@darling2008; @cote2016]. Multiple drivers can combine
non-linearly and result in effects that are greater (synergistic effect) or
lower (antagonistic effect) than the sum of individual effects
[@crain2008; @darling2008; @cote2016].

The uncertainty associated with complex driver interactions must therefore be
taken into account when investigating environmental impacts [@cote2016], yet
most research on driver effects in marine environments remains overwhelmingly
focused on single driver assessments [@obrien2019]. Increasing exposure and the
experiences of past ecological tragedies such as the collapse of cod fisheries
[@frank2005; @dempsey2018] and the decline of the beluga and right whale
populations [@plourde2014] together urge the need to characterize the distribution,
intensity and overlap between drivers in the St. Lawrence System. This will
provide critical information on areas most exposed to cumulative drivers and on
the interaction potential of drivers in the St. Lawrence. It is also a necessary
step towards the holistic and integrated management of the St. Lawrence System.

Gathering environmental data for large scale, systematic initiatives can,
however, be a very challenging and time consuming – not to say painful – process.
On one hand, there is an overwhelming and expanding wealth of data available.
Such information overload may inhibit our ability to make decisions based on
scientific information, promote massive effort duplication, disproportionately
appropriate research funds to certain sectors, and obscure knowledge gaps amid
a sea of information [@eppler2004]. On the other hand, crucial data are lacking
and remain largely unavailable or inaccessible for a variety of reasons,
including proprietary rights, lack of organizational time, capacity and
training, and in some rare cases unwillingness to share, curtailing our ability
for appropriate decision-making.

There are now initiatives that address this issue by assembling, organizing and
sharing environmental knowledge, such as the Ocean Biogeographic Information
System [OBIS; @obis2019] for biotic data and Bio-ORACLE [@tyberghein2012] for
abiotic data.
However, equivalent platforms for environmental drivers have largely focused on
single drivers (e.g. Global Fishing Watch) and platforms collating
data and knowledge on multiple drivers in a comparable and interoperable way
remain conspicuously missing [but see @halpern2015]. This is in spite of
integrated management and assessment approaches needing efficient data reporting,
standardized data management practices and tools tailored to the study of the
effects of multiple drivers [@dafforn2016; @stock2018]. An additional objective
thus emerged in the process of addressing our initial goals: sharing the
knowledge gathered through the description of drivers in the St.
Lawrence.

Our main objective in this study is to characterize the distribution and
intensity of drivers in the St. Lawrence. More specifically, we aim to: 1)
identify areas of high cumulative exposure, 2) identify drivers that are likely
to interact in the St. Lawrence and 3) characterize areas with similar
cumulative exposure regimes.
Here we focus solely on drivers, one of four elements required to evaluate
environmental impacts, *i.e.* drivers, ecosystem components of interest (*e.g.*
habitats or species) and the exposure and the vulnerability of the ecosystem
components to drivers. We achieve these objectives with the development of an
open knowledge platform named *eDrivers*. The platform was designed to
facilitate collaboration, real-time assessments of cumulative exposure and to
evolve with the addition of information and threats to the St-Lawrence ecosystems.

@@

## Estuary and Gulf of St. Lawrence

The St. Lawrence System is formed by the Estuary and the Gulf of St. Lawrence
(Figure \ref{egsl}). The Estuary is defined by the seawater limit
close to Îles d'Orléans to the west and by its connexion to the Gulf
near Pointe-des-Monts.
Estuarine circulation characterizes the Estuary. The surface layer is
composed of a freshwater seaward outflow primarily from the Great Lakes Basin
and through the St. Lawrence River, upstream of the Estuary. Atlantic waters are
flowing mainly from the Cabot Strait landward through the Laurentian Channel
(see below).

The Gulf is an interior sea connected to the Atlantic by the Cabot and Belle-Isle
Straits to the south and north of Newfoundland, respectively.
The topology of the Northern Gulf is characterized by three deep channels (250-500 m).
The Laurentian Channel is the main channel connecting the Estuary to the
Atlantic through the Cabot Strait. The Esquiman and Anticosti Channels are two
secondary channels that branch off from the Laurentian Channel to the north
towards the Arctic and the north of Anticosti Island, respectively.
The Southern Gulf is characterized by the Magdalen Shallows, a vast area with an
average depth of ~50 m. The Gulf is composed of a seasonal cold intermediate layer
that separates the surface and deep layers. Seasonal sea ice affects the circulation
of the St. Lawrence. Finally, three islands impact the physical dynamics of the
St. Lawrence: the Anticosti Island to the north, the Magdalen Islands in the
middle of the Magdalen Shallows and Prince Edward Island to the south. See
@saucier2003 and @galbraith2018 for more information on the physical oceanography
of the St. Lawrence.

The St. Lawrence drains over 25% of global freshwater reserves through its
connexion to the Great Lakes Basin. The Basin is home to over 45 million
North-Americans, *i.e.* 15 and 30 million in Canada and the United States,
respectively [@archambault2017]. The St. Lawrence itself boasts a much lower
population of approximately 1 million Canadians living within 10 km of the
coast, with populations mainly located in a few coastal cities in the Estuary
and the Southern Gulf [@statistics-canada2017].

@@

The list of drivers for which we sought data was informed by a global
cumulative impact assessment initiative [@halpern2008a; @halpern2015a],
regional holistic evaluations of the state of the St. Lawrence [@dufour2007;
@benoit2012], and communications with regional experts. We integrated datasets
from regional experts and also use global data from the global cumulative
impact assessment initiative [@halpern2008a; @halpern2015a] available from the
National Center for Ecological Analysis and Synthesis (NCEAS) online data
repository [Table 1; @halpern2015]. We selected global data that were
unavailable at the regional scale and that were available at a resolution
adequate for use at the scale of the St. Lawrence (*e.g.* marine pollution).

We characterized the intensity and distribution of 22 drivers
(Table 1;). Drivers incorporated in the analyses are varied in origin,
*i.e.* from terrestrial (*e.g.* nutrient input) to marine (*e.g.* shipping),
and from large scale biophysical processes (*e.g.* temperature anomalies) to
localized anthropogenic activities (*e.g.* fisheries). Drivers were divided
into 4 groups: coastal, climate, fisheries and marine traffic (Table 1). All
data layers and methodologies are described in the supplementary materials.

Drivers with non-normal frequency distributions were log-transformed (Figure S1)
and all drivers were scaled between 0 and 1 to allow driver comparisons. The
99th quantile of individual driver distribution was used as the upper bound for
scaling to control for extreme values and produce maps of individual drivers
(Figure S2). All drivers were embedded in a regular grid composed of 245604
$1 km^2$ hexagonal cells to construct the integrated dataset used for the
analyses.

@@

Areas with high cumulative exposure (objective 1) were identified by comparing
areas on the basis of the number and relative intensity of drivers in each
grid cell. Throughout the text, we use exposure when describing areas
exposed to drivers and we use footprint when describing the distribution and
intensity of drivers.

Cumulative footprint ($F$) was defined as the sum of the scaled intensity of all
drivers in each grid cell:

@@

where $x$ is a grid cell, $i$ is a driver and $D$ is the scaled intensity of
driver $i$. The cumulative footprint provides an estimate of the total relative
footprint in each grid cell. A grid cell with a high $F$ value is either
characterized by multiple drivers at low relative intensity, limited drivers at
high relative intensity, or both.

We also identified cumulative hotspots ($H$) to explore the the distribution of
cumulative exposure in the St. Lawrence (objective 1). Cumulative hotspots ($H$)
were defined as the number of drivers in each grid cell with scaled intensity
contained over their respective 80th percentile:

@@

Interactions among drivers were investigated using the cumulative footprint
($F$) between pairs of drivers (objective 2). The intensity at which pairs of
drivers co-occur was evaluated using a two-dimensional kernel density estimate.
As there are 231 pairwise combinations between 22 drivers, we focus on a single
example using hypoxia and demersal destructive fisheries, two drivers known to
occur mainly in deeper areas of the St. Lawrence and, hence, an interaction
between the effects of the two drivers could be anticipated.

@@

In order to better capture potential interactions between drivers (objective 2)
and to identify areas with similar exposure regimes (objective 3),
we identify threat complexes using a clustering approach [*e.g.* see @bowler2019].
We use the term clusters in presenting the methods, but use threat complex when
discussing the results on cumulative exposure regimes.

@@

Threat complexes were identified using a partional *k-medoids* clustering
algorithm, CLARA [CLustering for Large Applications; @kaufman1990], which was
designed for large datasets. The CLARA algorithm uses the PAM (Partition
Around Medoids) algorithm on a sample from the original dataset to identify a
set of $k$ objects that are representative of all other objects, *i.e.*
medoids and that are central to the cluster they represent. The goal of the
algorithm is to iteratively minimize intra-cluster dissimilarity. Iterations
are compared on the basis of the average dissimilarity between cluster objects
and representative medoid to select the optimal set of $k$ medoids that
minimizes average dissimilarity. We used the clustering algorithm with the
Manhattan distance since this measure is less affected by extreme values
[@legendre2012], as is the *k-medoids* clustering algorithm [@kaufman1990].
We used 100 iterations using samples of 10000 observations
(*i.e.* ~5% of observations) to identify clusters. Analyses were performed
using the *cluster* R package [@maechler2018].
Partitional clustering algorithms require a user-defined number of clusters.
Values of $k$ ranging from 2 to 10 were tested and validated by selecting
the number of clusters that maximized the average silhouette width [@kaufman1990]
and minimized the total within-cluster sum of squares (Figure S4).

@@

The difference between clusters was explored by measuring the total inter-cluster
dissimilarity and the contribution of each driver to the total inter-cluster
dissimilarity using a similarity percentage analysis (SIMPER) with Manhattan
distance [Figure S5; @clarke1993]. The Manhattan distance was again preferred for continuity
with the clustering analysis and to ensure that outliers did not have a strong
influence the analysis. As the drivers dataset is large (~ 250000 observations),
we used a bootstrap procedure for the SIMPER analysis, randomly selecting
5% of each cluster to run the analysis and repeating the process over 300
iterations. We also compared the mean intensity of each driver within each
cluster to better capture the inter-cluster dissimilarity.

@@

Intra-cluster similarity was evaluated calculating the intra-cluster Manhattan
distance and by transforming the mean contribution to distance ($M_c$) of each
driver by $.1 / (.1 + M_c)$ to obtain a similarity measure for each driver
($S_d$). The total similarity is the sum of all $S_d$. (Figure S6).
As with the inter-cluster dissimilarity, we used a bootstrap procedure for the
intra-cluster similarity, randomly selecting 25% of each cluster observation to
run the analysis and repeating the process over 50 iterations. We only used the
bootstrapping procedure for clusters with less than 40000 observations since
computation time was manageable.

<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->
# Results and Discussion
<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->

@@

Apart from the northeastern Gulf, the cumulative footprint of drivers is
ubiquitous in the St. Lawrence (Figure \ref{footprint}). Cumulative exposure is
generally highest along the coast (Figure \ref{footprint}), with hotspots
located in the vicinity of coastal cities (Figure \ref{hotspot}). In general,
offshore areas are less exposed to cumulative drivers, with the Estuary and the
Anticosti Gyre being notable exceptions (Figures \ref{footprint} and
\ref{hotspot}). This is not to say that offshore areas are free of exposure,
as most of the St. Lawrence is exposed to multiple overlapping drivers
(Figures \ref{footprint} and \ref{hotspot}). For example, it is worthy to note
high cumulative footprint observed at the heads of the Anticosti and Esquiman
Channels (Figure \ref{footprint}).

These results are consistent with observations elsewhere in the world, where
cumulative driver exposure conspicuously arises from and markedly intensifies
close to coastal cities and at the mouth of rivers draining highly populated
areas [e.g. @halpern2015a; @feist2016; @mach2017; @stock2018]. These are areas
where human activities (*e.g.* coastal development and shipping) and footprint
(*e.g.* pollution runoff) are the most intense [@feist2016], and on which is
overlaid a background of natural disturbances [@micheli2016]. They are also the
areas in which the most dramatic increases in exposure are expected, with
populations increasing more rapidly along the coast than inland [@feist2016].
In the St. Lawrence, large coastal cities are mostly located along the Estuary
and the southwestern Gulf, while the northeastern Gulf is largely uninhabited
or home to small coastal communities. Certain smaller coastal communities with
high cumulative footprint are characterized by large industries (*e.g.*
Sept-Îles and Charlottetown).

As for offshore exposure, the Estuary, along with the St. Lawrence River,
provide access to and serve as the primary drainage outflow of the Great Lakes
Basin, which is home to over 45 million North-Americans and is the most densely
populated region in Canada [@archambault2017; @statistics-canada2017]. Most marine
traffic thus converges to the Estuary.

@@

Box 1. Code snippet demonstrating how to use the *eDrivers* R package to
reproduce figure \ref{kernel}.
